---
title: "GWAS preprocessing for _seismic_ input"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GWAS preprocessing for seismic input}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The _seismic_ framework takes the output from [MAGMA](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1004219) as input. 
Basically, the MAGMA software digests SNP-level summary statistics of a trait in a genome-wide association
study into a gene-level MAGMA z-score statistics. This tutorial will guide you through the
preprocessing steps necessary to convert raw GWAS summary statistics into the proper format
for _seismic_ analysis.

Before beginning, ensure you have installed the MAGMA software and obtained the required files:

- Gene location file
- SNP synonyms file (dbSNP)
- Population-specific reference data

These files can be all downloaded from the [MAGMA website](https://cncr.nl/research/magma/). 

In addition to these reference files, a GWAS summary statistics table is also required to provide
SNP-level information for a trait of interest. We will use example files to demonstrate the preprocessing
steps. These files are available for download from our data repository [here](https://zenodo.org/records/15582078).
We will assume that users wanting to follow this tutorial have downloaded and unzipped the `all_data` folder
in this repository before beginning.

## Step 1: Clean up the GWAS summary statistics file

A starting GWAS summary statistics file should at minimum include columns for: chromsome ID,
base-pair position, and p-value.

Additionally, we strongly recommend including SNP IDs to facilitate MAGMA analysis.

```{r show-only, eval=FALSE, echo=TRUE}
library(magrittr)
library(dplyr)

example_gwas <- read.table("all_data/example_data/sample_gwas.txt", header = T) %>% as_tibble()

head(example_gwas)
```

```console
## # A tibble: 6 × 7
##   Allele1 Allele2 Direction P.value  group Chromosome     BP
##   <chr>   <chr>   <chr>       <dbl>  <int>      <int>  <int>
## 1 t       c       ---        0.0541 184129          1 796338
## 2 a       g       +++        0.0315  46089          1 817341
## 3 t       c       +++        0.0159 189219          1 817514
## 4 a       t       -+-        0.199   39112          1 870176
## 5 c       g       --+        0.730   80434          1 928622
## 6 t       c       --+        0.913  185512          1 996120
```

## Step 2: Map the SNP location to GRCh37 

For MAGMA gene analysis, the required population-specific reference data is based on the GRCh37 genome build,
all SNP location should be converted to GRCh37 (hg19). We illustrate the steps of this conversion using
[LiftOver](https://genome.ucsc.edu/cgi-bin/hgLiftOver) provided by UCSC. We start with hg38 data, but if
your data is already hg19 you can skip this step. Additionally, this preprocessed version is already 
saved in the downloadable example files: `all_data/example_data/sample_gwas_hg19.txt`.

First the chain object `hg38ToHg19.over.chain.gz`
must be downloaded from the [UCSC website](https://hgdownload.cse.ucsc.edu/goldenpath/hg19/liftOver/)
and unzipped. 

```{r show-only, eval=FALSE, echo=TRUE}
library('GenomicRanges')
library('IRanges')
library('rtracklayer')

chainObject <- import.chain("hg38ToHg19.over.chain") # update with path to chain obj
grObject <- GRanges(seqnames = paste0("chr", example_gwas$Chromosome),
                   ranges = IRanges(start=as.numeric(example_gwas$BP), end=as.numeric(example_gwas$BP)+1))

hg19_res <- liftOver(grObject, chainObject)

# extract the result data frame, select the necessary columns
hg19_res_df <- as.data.frame(hg19_res) %>%
  mutate(seqnames = gsub(seqnames, pattern = "chr",replacement = "")) %>%
  select(group, start, seqnames) %>%
  rename("start" = "BP", "seqnames" = "Chromosome")

# get the 1 to 1 mapping
hg19_res_df <- hg19_res_df %>% group_by(group) %>% 
  filter(n()==1) %>%
  ungroup() 

# merge with the original data frame (the group column indicates the original position in the table)
example_gwas <- example_gwas %>%
  mutate(group = 1:n()) %>%
  select(-Chromosome, -BP) %>%
  left_join(hg19_res_df , by = "group") %>%
  filter(!is.na(BP))

write.table(example_gwas %>% mutate(BP = as.character(BP)),  #avoid numerical output 
            file = "sample_gwas_hg19.txt", 
            sep="\t", col.names = T, row.names = F, quote = F)

head(example_gwas)
```

## Step 3: Map the SNP location to SNP ID

If the SNP ID is not included in the GWAS summary statistics file, we can use the
SNP synonyms file to map the SNP location to SNP ID. 
To simplify the procedure, we have included a bash script in the [seismic-analysis repository](https://github.com/ylaboratory/seismic-analysis)
at `src/tools/annotate_SNP.sh`. 
This script will require the dbSNP file and the GWAS summary statistics file as input, and output a new file with SNP ID included in the last column.
To view all specific parameters of the script, you can run the following command in the terminal: 

```bash
bash src/tools/annotate_SNP.sh -h
```

```console
## 
##     This script help you locate SNP and transform the locus (by chromosome id and location) to the corresponding variant rsid.
##     Only intersected SNPs will be printed out.
##     Note!!!! 1. All file path should be the same as the original files. 2. bedtools is required, so make sure it's in your environment or path
##     Here are the required parameters:
##     -g | GWAS summary statistics files. Or any other files that you would like to annotate SNPs to.
##     -o | Outpur file header. If the value is given as 'dir1/dir2/output', then the final output file will be 'dir1/dir2/output.annot.table.txt'. If it is not specified, the output file will be the same as the input file with the suffix ignored.
##     -c | Chromosome column, i.e. the column with with chromosome information be (in the table). The value should be specified or by default it will be 1.
##     -p | Position column, i.e. The column with the the position information (in the table), The value should be specified or by default it will be 2.
##     -b | Basis of chromosome numbering, to specify the if first position on a chromosome is labeled as '1'(1-base), or '0'(0-base). The default parameters are start with 1. #only these two are available!
##     -k | Keep all unmapped SNPs which do not exist in the dbSNP file or not. By default it is True. (You can choose between 'True' or 'False')
##     
##     optional parameters:
##     -v | VCF file for the dbsnp version you used. For default it will be located in 'data/ref/Hg19.dbsnp151.vcf'. This can be downloaded online.
##     -n | Number of tasks to split the job. The default value is 1. If the memory is not enough, you can split the job into several parts but it will take more time.
##     -t | Customized bedtools command. By default it is 'bedtools'. If you do not have it in your environment, you can specify the path to the bedtools command.
##     -w | The working directory. By default it is the current directory. If you want to specify the working directory, you can use this parameter.
##     -h | Help message.
## 
```


To annotate locations, this script requires a vcf file for the dbSNP, which
is available for download from UCSC, and requires
[bedtools](https://bedtools.readthedocs.io/en/latest/). Once these prerequisties are met
you can run the script as follows:

```bash
bash src/tools/annotate_SNP.sh -g all_data/example_data/sample_gwas_hg19.txt -o all_data/example_data/sample_gwas_hg19 \
-c 6 -p 7 -v data/ref/Hg19.dbsnp151.vcf -t bin/bedtools
```

Note that the script may takes hours to run and consumes more than 100 GB of memory.
To alleviate memory usage but increase the running time, the `-n` option may be used
to sequentially split the task. 

The resulting annotation table will be printed with the path and header specified by the `-o` option. 

```console
> head all_data/example_data/sample_gwas_hg19.annot.table

## Allele1  Allele2 Direction   P.value group   Chromosome  BP  marker.name
## t    c   --- 0.0541  1   1   731718 rs58276399
## a    g   +++ 0.03149 2   1   752721 rs3131972
## t    c   +++ 0.01587 3   1   752894 rs3131971
## a    t   -+- 0.1987  4   1   805556 rs72631880
## c    g   --+ 0.7303  5   1   864002 rs1806501
## t    c   --+ 0.913   6   1   931500 rs74718486
## a    g   --+ 0.4902  7   1   938125 rs2710868
## a    g   --+ 0.785   8   1   941334 rs57683598
## t    c   +++ 0.1863  9   1   948921 rs15842
```

## Step 4: Generate the input file for MAGMA

The MAGMA software takes two input files generated from the previously processed
(or unprocessed but with the required columns) GWAS summary statistics file:
a SNP location file and a p-value file.
You can simply use the `awk` command to extract the information in the correct format
in the command line:

```bash
# print rsID, chromosome, BP but not colum names:
awk '{if (NR>1) print $8, $6, $7}' all_data/example_data/sample_gwas_hg19.annot.table > all_data/example_data/sample_gwas_hg19.snp_loc.txt
# print rsID, p-value but not colum names:
awk '{if (NR>1) print $8, $4}' all_data/example_data/sample_gwas_hg19.annot.table > all_data/example_data/sample_gwas_hg19.p_val.txt
```

or in R:

```R
anno_tbl <- read.table("all_data/example_data/sample_gwas_hg19.annot.table", header=T)
anno_tbl %>% select(marker.name, Chromosome, BP) %>% write.table(file = "all_data/example_data/sample_gwas_hg19.snp_loc.txt", col.names = F, row.names = F, quote = F, sep="\t")
anno_tbl %>% select(marker.name, P.value) %>% write.table(file = "all_data/example_data/sample_gwas_hg19.p_val.txt", col.names = F, row.names = F, quote = F, sep="\t")
```

This results in a file that looks like:

```console
> head all_data/example_data/sample_gwas_hg19.snp_loc.txt

## rs58276399   1   731718
## rs3131972    1   752721
## rs3131971    1   752894
## rs72631880   1   805556
## rs1806501    1   864002
## rs74718486   1   931500
## rs2710868    1   938125
## rs57683598   1   941334
## rs15842  1   948921
## rs9442364    1   970215
```

## Step 5: Run MAGMA software

There are two steps: gene annotation and gene analysis.
You can use our implemented bash script `src/tools/magma_gene_zscore_analysis.sh` in the
[seismic-analysis repository](https://github.com/ylaboratory/seismic-analysis) 
to run the two steps together.

Another parameter to specify is the gene analysis window size, which is set
as 35 kilobase (kb) upstream to 10 kb downstream by default, as recommend
by [a previous study](https://www.nature.com/articles/s41588-020-0610-9).
Although we have shown that output files from various window sizes make only
a minor difference in the results of _seismic_ analysis, you can still change
the window size by specifying the parameter `--window-size` in the script. 

To view all specific parameters of the script, you can run the following command in the terminal:

```console
> bash src/tools/magma_gene_zscore_analysis.sh -h

## Usage: magma_gene_zscore_analysis.sh -l SNP_LOC_FILE -p SNP_P_VALUE_FILE -o OUTPUT_DIR -s COHORT_SIZE -m MAGMA_PATH -g GENE_LOC_FILE -b B_FILE [-w WINDOW_SIZES] [-h]
## 
## This script performs MAGMA annotation and analysis on SNP data.
## 
## Required parameters:
##     -l | SNP location file
##     -p | SNP p-value file
##     -o | Output directory
##     -s | Cohort size (use 'cX' format to specify column X containing sample size)
##     -m | Path to MAGMA binary
##     -g | Path to gene location file
##     -b | Path to g1000_eur files (without extension)
## 
## Optional parameters:
##     -w | Window sizes (default: 35,10). Use colon-separated list for multiple sizes, e.g., '35,10:15,20:40,10'
##     -h | Display this help message
```

Then finally to run MAGMA with its accompanying [reference data](https://cncr.nl/research/magma/):

```bash
bash src/tools/magma_gene_zscore_analysis.sh -l all_data/example_data/sample_gwas_hg19.snp_loc.txt \
-p all_data/example_data/sample_gwas_hg19.p_val.txt \
-o all_data/example_data/ \
-s 100000 -m bin/magma/magma -g ref/magma/NCBI37.3/NCBI37.3.gene.loc -b ref/magma/g1000_eur/g1000_eur 
```
Besides specifying these reference file paths and the curated files from the summary statistics,
the total size of the cohort should also be specified. Note that sometimes the cohort size for
each SNP may be different, for example, due to some quality control procedure. In that case, a
separate column in the p-value file should be added and the input `-s` argument should tell the
script the order of the column (for example, the third column) describing the cohort size. By
specifying these arguments, the gene-level MAGMA z-score will be printed out in the output
directory with header similar to the snp location file.
In our example the output file will be: `all_data/example_data/sample_gwas_hg19.35.10.genes.out`.
In addition, please remember to check the MAGMA running log. Sometimes when 
the data are not in the correct format, there will be very few genes
and SNPs annotated (e.g., less than 10% of the SNPs).
